Reproducible Research: Peer Assessment 1 (Nov. 2014)
---


## Global options and parameters
```{r setglobals}
#setwd("~/Documents/Coursera/datascience_jh/represearch/RepData_PeerAssessment1")
# set echo = TRUE globally. This is the default, setting it explicitly here
opts_chunk$set(echo = TRUE)

# disable scientific notation
options(scipen = 9)
# we will be using the packages knitr to convert to markdown and html
# Using knit2html will generate the .md and .html files in the same directory
# as the working directory. A directory called figures will also be created 
# under the working directory if plots are generated.
library(knitr)
# load ggplot2 package for plots
library(ggplot2)
```


## Loading and preprocessing the data
```{r loading}
if (!file.exists('activity.csv')){
    unzip("activity.zip")
}
activity <- read.csv("activity.csv", header = TRUE)
```


## What is mean total number of steps taken per day?
```{r meanTotalStepsPerDay}
# calculate total number of steps grouped by day
totstepsbyday <- aggregate(steps~date, activity, sum, na.rm = TRUE)
# Q1 - Make a histogram of the total number of steps taken each day

hist(totstepsbyday$steps, col = "orange", main = "Total Steps taken each day",
      xlab = "Total number of steps taken each day")
# Q2 - Calculate and report the mean and median total number of steps taken per day
meanstepsbyday <- round(mean(totstepsbyday$steps))
medianstepsbyday <- median(totstepsbyday$steps)
```
#### The mean and median total number of steps taken per day are `r meanstepsbyday` and `r medianstepsbyday`


## What is the average daily activity pattern?
```{r avgDailyActivityPattern}
# Q1 - Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) 
# and the average number of steps taken, averaged across all days (y-axis)

# calculate the average number of steps grouped by interval
avgstepsbyinterval <- aggregate(steps~interval, activity, mean)
# generate line chart
plot(avgstepsbyinterval$interval, avgstepsbyinterval$steps, type = 'l', col = 1, 
     main = "Average number of steps across all days",
     xlab = "Interval in minutes",
     ylab = "Average number of steps")

# Q2 - Which 5-minute interval, on average across all the days in the dataset, 
# contains the maximum number of steps?

# find the row that has the maximum number of steps
rowWithMax <- which.max(avgstepsbyinterval$steps)
# get the values for the interval and steps corresponding to the rowWithMax row
avgstepsbyinterval[rowWithMax, ]
```
#### The interval `r avgstepsbyinterval[rowWithMax, ]$interval` corresponds to the maximum average number of steps  `r avgstepsbyinterval[rowWithMax, ]$steps`


## Imputing missing values
```{r fillMissingValues}
# Q1. Calculate and report the total number of missing values in the dataset 
# (i.e. the total number of rows with NAs)

# subset activity data frame to get rows with steps = NA
onlyNA <- subset(activity, is.na(steps))
rowcountNA <- nrow(onlyNA)
```
#### The total number of missing values in the dataset is `r rowcountNA`.
```{r}
# Q2 - Devise a strategy for filling in all of the missing values in the dataset. 
# The strategy does not need to be sophisticated. For example, you could use the
# mean/median for that day, or the mean for that 5-minute interval, etc.
# use sqldf for getting and setting missing values
# use the mean value for an interval to substitute for NAs

library(sqldf)

# create a dataframe that combines the avgstepsbyinterval data frame with the 
# onlyNA data frame
interimdf <- sqldf("select round(a.steps) a_st, a.interval a_int, na.steps na_st,na.interval na_int,
na.date from avgstepsbyinterval a join onlyNA na on a.interval = na.interval")

# create a new activity dataframe with NAs filled in
activityfilled <- sqldf("select a.date, a.interval, coalesce(t.a_st,a.steps) steps
                        from activity a left outer join tst t 
                        on (a.date = t.date and a.interval = t.na_int)")

# calculate total number of steps grouped by day
totstepsbydayfill <- aggregate(steps~date, activityfilled, sum)

# create histogram
hist(totstepsbydayfill$steps, col = "blue", main = "Total Steps taken each day",
      xlab = "Total number of steps taken each day")
meanstepsbydayfill <- round(mean(totstepsbydayfill$steps))
meanstepsbydayfill
medianstepsbydayfill <- median(totstepsbydayfill$steps)
medianstepsbydayfill
```
#### The mean for the dataset with NAs present is `r meanstepsbyday`.
#### The mean for the dataset with NAs filled in is `r meanstepsbydayfill`.
#### The median for the dataset with NAs present is `r medianstepsbyday`.
#### The median for the dataset with NAs filled in is `r medianstepsbydayfill`.
#### Imputing missing data has no difference on the mean but has a slight impact on the median.


## Are there differences in activity patterns between weekdays and weekends?
```{r activityPatterns}
# Q1 - Create a new factor variable in the dataset with two levels – “weekday” 
# and “weekend” indicating whether a given date is a weekday or weekend day.

# convert the data factor variable into dates
activityfilled$date <- as.Date(activityfilled$date, "%Y-%m-%d")

# add new column to dataframe to indicate type of day
# The wday component of a POSIXlt object is the numeric weekday 
# (0-6 starting on Sunday).
# In order to match up with the character vector add 1 to the value
# returned by wday. The returned value will be a char datatype
activityfilled$daytype <- c("Weekend", "Weekday", "Weekday", "Weekday", "Weekday", 
  "Weekday", "Weekend")[as.POSIXlt(activityfilled$date)$wday + 1]

# convert to factor
activityfilled[ , "daytype"] <- as.factor(activityfilled[, "daytype"])
str(activityfilled)

# Q2 - Make a panel plot containing a time series plot (i.e. type = "l") of 
# the 5-minute interval (x-axis) and the average number of steps taken, 
# averaged across all weekday days or weekend days (y-axis).
avgstepsbyintday <- aggregate(steps ~ interval + daytype , activityfilled, mean)
qplot(interval, steps, data = avgstepsbyintday, color = daytype, geom = c("line"),
      xlab = "Interval in Minutes", ylab = "Average Number of steps",
      main = "") + facet_wrap(~ daytype, ncol = 1)
```

